### 2025-05-05 Chapters/Chapter1.tex
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

This thesis examines methods to predict the gust factor using various predictors and multiple data sources, including NWP outputs and observational data. Accurate gust predictions are crucial, as it is ofteh the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines. Moreover, the prevalence of extreme wind events is expected to increase in the future \cite{nasa_extreme_weather}.

@@

\subsection*{Numerical weather prediction}
The history of numerical weather prediction dates back to the 1920s, when Lewis Fry Richardson pioneered the field and attempted to produce forecasts. His results were flawed due to numerical noise. The ENIAC, built in 1945, was a general-purpose computer used—among other tasks—for weather prediction. These forecasts took 24 hours to compute and predicted 24 hours into the future. While a proof of concept, they were not operationally useful \cite{TheENIACForecastsARecreation}.

@@

\subsection*{AI in weather forecasts}
In the last decade, there has been another transformation in weather prediction driven by artificial intelligence (AI). Interest in AI has come in waves: progress is made, then interest diminishes, but over the past 15 years growth has been steady. Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the developement of neural networks (NNs) for processing massive datasets, and the availability of large online datasets. One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \cite{canNNBeatNWP}. Since 2018, significant work has applied AI to weather prediction.

@@

\subsection*{Large weather models}
In the last two years, there have been even more developments with the emergence of Large AI Weather forecast Models (LWMs). In 2024, Ling et al.\ \cite{SecondRevolution} proposed a standardized definition of LWMs in meteorology, outlining three criteria, referred to as the "Three Large Rules":

@@

Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace tradional NWP systems. In early 2022, Pathak et al.\ \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms them for smaller-scale features. It generates a one-week forecast in under two seconds—orders of magnitude faster than conventional physical methods \cite{FourCastNet}. In 2022, several ML-based models demonstrated faster-than-NWP predictions after one-time training, with performance rivaling or exceeding NWP in some cases.

@@

\subsection*{Conclusion}
Substantial progress has been made since 2018 and especially since 2022 \cite{SecondRevolution}. The evolution from proof-of-concept ML methods to models competitive with—or surpassing—standard NWP has been remarkably rapid. It is important to note that training data for these large models are drawn from traditional NWP outputs, underscoring the synergy between ML and physics-based approaches. The coming years promise further exciting developments in machine-learning–based weather prediction.

\section{Methodology and related work}
In this study, data from three sources is used:

\begin{enumerate}[label=\textbf{\arabic*}.,rightmargin=1.5em]
  \item Three-hourly reanalysis data for Iceland (2004–2023)
  \item Dense (20 by 20~m) gridded elevation data for Iceland
  \item Three-hourly observations from Icelandic meteorological stations
\end{enumerate}

@@

These activation control neural activation and can help stabilize the network.
\subsection{Model evaluation}
To measure the performance of these models, both in training and testing, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
\begin{equation}
    \label{eqn:mape}
    \text{MAPE} = \frac{1}{n}\sum_{i=1}^n\frac{|y_{\mathrm{predict}} - y_{\mathrm{true}}|}{y_{\mathrm{predict}}}
\end{equation}
This measure was chosen because the target is the gust factor (the wind gust over the average wind). If the target had been the wind gust rather than the gust factor, mean absolute error might be more appropriate.

>>>>

This thesis examines methods to predict the gust factor using various predictors and multiple data sources, including NWP outputs and observational data. Accurate gust predictions are crucial, as it is often the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines. Moreover, the prevalence of extreme wind events is expected to increase in the future \cite{nasa_extreme_weather}.

@@

\subsection{Numerical weather prediction}
The history of numerical weather prediction dates back to the 1920s, when Lewis Fry Richardson pioneered the field and attempted to produce forecasts. His results were flawed due to numerical noise. The ENIAC, built in 1945, was a general-purpose computer used—among other tasks—for weather prediction. These forecasts took 24 hours to compute and predicted 24 hours into the future. While a proof of concept, they were not operationally useful \cite{TheENIACForecastsARecreation}.

@@

\subsection*{AI in weather forecasts}
In the last decade, there has been another transformation in weather prediction driven by artificial intelligence (AI). Interest in AI has come in waves: progress is made, then interest diminishes, but over the past 15 years growth has been steady. Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the development of neural networks (NNs) for processing massive datasets, and the availability of large online datasets. One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \cite{canNNBeatNWP}. Since 2018, significant work has applied AI to weather prediction.

@@

\subsection{Large weather models}
In the last two years, there have been even more developments with the emergence of Large AI Weather forecast Models (LWMs). In 2024, Ling et al.\ \cite{SecondRevolution} proposed a standardized definition of LWMs in meteorology, outlining three criteria, referred to as the "Three Large Rules":

@@

Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace traditional NWP systems. In early 2022, Pathak et al.\ \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms them for smaller-scale features. It generates a one-week forecast in under two seconds—orders of magnitude faster than conventional physical methods \cite{FourCastNet}. In 2022, several ML-based models demonstrated faster-than-NWP predictions after one-time training, with performance rivaling or exceeding NWP in some cases.

@@

\subsection{Conclusion}
Substantial progress has been made since 2018 and especially since 2022 \cite{SecondRevolution}. The evolution from proof-of-concept ML methods to models competitive with—or surpassing—standard NWP has been remarkably rapid. It is important to note that training data for these large models are drawn from traditional NWP outputs, underscoring the synergy between ML and physics-based approaches. The coming years promise further exciting developments in machine-learning–based weather prediction.

\section{Methodology and related work}
In this study, data from three sources are used:

\begin{enumerate}[label=\textbf{\arabic*}.,rightmargin=1.5em]
  \item Three-hourly reanalysis data for Iceland (2004–2023)
  \item Dense (20 by 20~m) gridded elevation data for Iceland
  \item Hourly observations from Icelandic meteorological stations
\end{enumerate}

@@

These activation functions control neural activation and can help stabilize the network.
\subsection{Model evaluation}
To measure the performance of these models, both in training and testing, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
\begin{equation}
    \label{eqn:mape}
    \text{MAPE} = \frac{1}{n}\sum_{i=1}^n\frac{|y_{\mathrm{predict}} - y_{\mathrm{true}}|}{y_{\mathrm{predict}}}
\end{equation}
This measure was chosen because the target is the gust factor (the wind gust over the average wind). If the target had been the wind gust rather than the gust factor, mean absolute error might be more appropriate.

### 2025-05-05 Chapters/Chapter2.tex
Update Chapter2.tex
Updated images of flow chart and station distribution over iceland
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

IMO provided measurements from \nStationsMin weather stations across Iceland. The measurements that met the filtering criteria began in \startDateVedur and ended in 2023. Of these \nStationsMin stations, \nVedurMin were from the IMO, with the anemometer at 10~m above ground, while the remaining \nVGMin stations were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)}, with the anemometer at 6–7~m above ground \cite{vegagerdin_postur}. The locations of these weather stations are shown in Figure \ref{fig:aws_map}.

Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases. Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.

@@

\begin{figure}
    \centering
    \includegraphics[scale = 1]{Figures/stationsOverIceland_2024-05-16_stripped_to_frame.png}
    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study. Most of these were from IMO but over a hundred were from IRCA. IMO anemometers are placed at 10 meters above ground, while IRCA ones are placed at 6--7 meters above ground.}
    \label{fig:aws_map}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{Figures/data-preprocessing-flow-chart.png}
  \caption{Flow chart illustrating the data-combination workflow}
  \label{fig:data_preprocessing_flow_chart}
\end{figure}

>>>>

Automatic weather stations (AWS) have been set up all around the world, including Iceland. AWSs might include a variety of sensors, such as barometers, thermometers, and anemometer along with solar panels (or other power source) and some kind of data logger. IMO provided measurements from AWSs across Iceland. One of the stations used in this study can be seen in Figure (\ref{fig:storhofdi}). Not every station had consistently available data. Data from \nStationsMin AWSs were included looking at a period from \startDateVedur and ended in 2023. Of these \nStationsMin stations provided by IMO, \nVedurMin were IMO stations, with the anemometer at 10~m above ground, while the remaining \nVGMin stations were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)}, with the anemometer at 6–7~m above ground \cite{vegagerdin_postur}. The locations of these weather stations are shown in Figure \ref{fig:aws_map}.

\begin{figure}
  \centering
  \includegraphics[scale=0.3]{Figures/storhofdi.jpeg}
  \caption[AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.]{AWS at Stórhöfði in Vestmannaeyjar, south of Iceland. The anemometer can be seen as three pins in the upper-central of the figure just slightly to the left.}
\end{figure}

Data from these AWS were stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases. Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.

@@

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/station_elevations.png}
    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study. Most of these were from IMO but over a hundred were from IRCA. IMO anemometers are placed at 10 meters above ground, while IRCA ones are placed at 6--7 meters above ground.}
    \label{fig:aws_map}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.6]{Figures/data_combination.drawio.png}
  \caption{Flow chart illustrating the data-combination workflow}
  \label{fig:data_preprocessing_flow_chart}
\end{figure}

### 2025-05-05 Chapters/Chapter3.tex
komið lengra
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

\chapter{Model Architecture and Training}

@@

\section{Model structure}
The network consists of \(n\) fully connected layers, with \(n\) around 10. Each hidden layer has the same number of units and is followed by batch normalization. A dropout layer with rate 50\% is applied after the final hidden layer. A dense output layer with one unit produces the predicted gust factor. A grid search was used to find the hyperparameters that minimize the loss. Hyperparameters are settings defined before training begins \cite{hyperparameters_definition}. They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty. The ranges tested are listed in Table \ref{table:gridSearchHyperparamters}.

\begin{table}[h]
    \centering
    \caption[Hyperparameter search with Hyperband]{Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones. The table shows the ranges tested and the best-performing combination. Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
    \label{table:gridSearchHyperparamters}
    \begin{tabular}{lll}
        \toprule
        Parameter & Range of values & Selected\\
        \midrule
        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{250}\\
        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
        Activation & ReLU, ELU, tanh & \textbf{ReLU}\\
        \bottomrule
    \end{tabular}
\end{table}

As mentioned, hyperband doesn't set a hard upper limit to the number of epochs it will train in total. When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{hyperband class} several factors can be set. One of interest here is the $hyperband_iterations$ argument. This determines how often the hyperband algorithm is run and defaults to 1. For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for. Searching a space generally takes a lot of time but this drastically improves on gridsearch. If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource. This is resource intensive and cannot be repeated often. Another question that remains is whether the ranges given are optimal.

As noted, Hyperband does not impose a fixed limit on the total number of training epochs. When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the parameter \texttt{hyperband\_iterations} controls how many times Hyperband is run and defaults to 1. On each run, the total number of epochs is distributed across the hyperparameter trials. The total number of epochs is approximately

\begin{equation*}
n_{\mathrm{epochs}} = \text{max\_epochs} \times \bigl(\log(\text{max\_epochs})\bigr)^2 \approx 10^4
\end{equation*}

for $ \text{max\_epochs} = 1000$. Although this method is more efficient than grid search, if each epoch takes about 10 seconds, a complete Hyperband search can still require about 28 hours on shared resources, making repeated searches impractical. It remains unclear whether the current hyperparameter ranges are optimal.

>>>>

\chapter{Model Architecture}

@@

The network consists of $n$ fully connected layers, with $n$ around 10. Each hidden layer has the same number of units and is followed by batch normalization. A dropout layer with rate 50\% is applied after the final hidden layer. A dense output layer with one unit produces the predicted gust factor. A grid search was used to find the hyperparameters that minimize the loss. Hyperparameters are settings defined before training begins \cite{hyperparameters_definition}. They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty. The ranges tested are listed in Table \ref{table:gridSearchHyperparameters}.

\begin{table}[h]
    \centering
    \caption[Hyperparameter search with Hyperband]{Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones. The table shows the ranges tested and the best-performing combination. Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
    \label{table:gridSearchHyperparameters}
    \begin{tabular}{lll}
        \toprule
        Parameter & Range of values & Selected\\
        \midrule
        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{250}\\
        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
        Activation & ReLU, ELU, tanh & \textbf{ReLU}\\
        \bottomrule
    \end{tabular}
\end{table}

As mentioned, Hyperband doesn't set a hard upper limit to the number of epochs it will train in total. When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband class} several factors can be set. One of interest here is the Hyperband iterations argument. This determines how often the Hyperband algorithm is run and defaults to 1. For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for. Searching a space generally takes a lot of time but this drastically improves on grid search. If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource. This is resource-intensive and cannot be repeated often. Another question that remains is whether the ranges given are optimal.

As noted, Hyperband does not impose a fixed limit on the total number of training epochs. When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the parameter \texttt{Hyperband\_iterations} controls how many times Hyperband is run and defaults to 1. On each run, the total number of epochs is distributed across the hyperparameter trials. The total number of epochs is approximately
%
\begin{equation*}
n_{\mathrm{epochs}} = \text{max\_epochs} \times \bigl(\log(\text{max\_epochs})\bigr)^2 \approx 10^4
\end{equation*}
%
for $ \text{max\_epochs} = 1000$. Although this method is more efficient than grid search, if each epoch takes about 10 seconds, a complete Hyperband search can still require about 28 hours on shared resources, making repeated searches impractical. It remains unclear whether the current hyperparameter ranges are optimal.
